---
title: "Marquee Integration for Colored Labels"
vignette: >
  %\VignetteIndexEntry{Marquee Integration for Colored Labels}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(stwd)
library(ggplot2)
library(marquee)
```

## Overview

The [marquee package](https://marquee.r-lib.org/) provides rich text
formatting for R graphics, including the ability to color text inline.
The stwd package integrates with marquee to create labels where the text
color automatically matches data values.
## Why Colored Labels?

In data visualizations, you often want labels to match their
meaning:

- **Green text** for positive variances
- **Red text** for negative variances

This creates visual consistency and immediate comprehension.

## The marquee_color_labels() Function

This function formats text labels with marquee color syntax:

```{r}
#| label: basic-example
# Basic usage
labels <- c("Good", "Bad", "Neutral")
colors <- c("#006400", "#8B0000", "#404040")

marquee_color_labels(labels, colors)
```

### With Formatting Options

```{r}
#| label: formatted
# Bold labels
marquee_color_labels(labels, colors, bold = TRUE)
```

```{r}
#| label: italic
# Italic labels
marquee_color_labels(labels, colors, italic = TRUE)
```

```{r}
#| label: bold-italic
# Bold and italic
marquee_color_labels(labels, colors, bold = TRUE, italic = TRUE)
```

## Using with ggplot2 and marquee

### Manual Approach

You can use `marquee_color_labels()` to pre-format your data, then use
`marquee::geom_marquee()`:
```{r}
#| label: manual-approach
#| fig-width: 8
#| fig-height: 5
#| eval: false
library(marquee)

#  Define colors for examples
cols <- list(actual = "#404040", variance_positive = "#006400", variance_negative = "#8B0000", previous = "#A0A0A0")

# Prepare data
variance_data <- data.frame(
    category = c("Product A", "Product B", "Product C", "Product D"),
    value = c(120, 85, 105, 95),
    variance = c(20, -5, 5, -5)
)

# Determine colors based on variance
variance_data$text_color <- ifelse(
    variance_data$variance >= 0,
    cols$variance_positive,
    cols$variance_negative
)

# Create formatted labels
variance_data$formatted_label <- marquee_color_labels(
    sprintf("%+d", variance_data$variance),
    variance_data$text_color,
    bold = TRUE
)

# Plot with marquee
ggplot(variance_data, aes(x = category, y = value)) +
    geom_col(fill = cols$actual, width = 0.6) +
    geom_marquee(
        aes(label = formatted_label),
        vjust = -0.5,
        size = 4
    ) +
    labs(title = "Sales with Colored Variance Labels") +
    theme_minimal()
```

### Using geom_marquee_colored()

For convenience, use `geom_marquee_colored()` which handles the formatting
automatically:

```{r}
#| label: geom-colored
#| fig-width: 8
#| fig-height: 5
#| eval: false
library(marquee)

variance_data$label <- sprintf("%+d", variance_data$variance)

ggplot(variance_data, aes(x = category, y = value)) +
    geom_col(fill = cols$actual, width = 0.6) +
    geom_marquee_colored(
        data = variance_data,
        mapping = aes(x = category, y = value + 5, label = label, color = text_color),
        bold = TRUE,
        size = 4
    ) +
    labs(title = "Sales with Auto-Colored Variance Labels") +
    theme_minimal()
```

## Practical Examples

### Variance Chart with Colored Labels

```{r}
#| label: variance-example
#| fig-width: 9
#| fig-height: 5
library(marquee)

#  Define colors for examples
cols <- list(actual = "#404040", variance_positive = "#006400", variance_negative = "#8B0000", previous = "#A0A0A0")

# Sales performance data
sales <- data.frame(
    region = c("North", "South", "East", "West", "Central"),
    actual = c(150, 120, 180, 95, 110),
    plan = c(140, 130, 170, 100, 100)
)

sales$variance <- sales$actual - sales$plan
sales$variance_pct <- round((sales$variance / sales$plan) * 100, 1)
sales$var_color <- ifelse(sales$variance >= 0,
                          cols$variance_positive,
                          cols$variance_negative)
sales$label <- sprintf("%+.1f%%", sales$variance_pct)

ggplot(sales, aes(x = reorder(region, variance))) +
    # Reference bars (plan)
    geom_col(aes(y = plan), fill = NA, color = cols$actual,
             linewidth = 0.8, width = 0.6) +
    # Actual bars
    geom_col(aes(y = actual), fill = cols$actual, width = 0.4) +
    # Colored variance labels
    geom_marquee_colored(
        data = sales,
        mapping = aes(
            x = reorder(region, variance),
            y = actual + 8,
            label = label,
            color = var_color
        ),
        bold = TRUE,
        size = 3.5
    ) +
    coord_flip() +
    labs(
        title = "Regional Sales Performance vs Plan",
        x = NULL, y = "Sales ($M)"
    ) +
    theme_minimal()
```

### Waterfall with Colored Value Labels

```{r}
#| label: waterfall-colored
#| fig-width: 10
#| fig-height: 5
library(marquee)

#  Define colors for examples
cols <- list(actual = "#404040", actual_light = "#808080", variance_positive = "#006400", variance_negative = "#8B0000", previous = "#A0A0A0")

# Bridge data
bridge <- data.frame(
    driver = c("2023", "Volume", "Price", "Mix", "FX", "2024"),
    value = c(100, 15, 8, -5, -3, NA),
    type = c("start", "positive", "positive", "negative", "negative", "end")
)

# Calculate positions for waterfall
bridge$running <- cumsum(ifelse(is.na(bridge$value), 0, bridge$value))
bridge$running[nrow(bridge)] <- bridge$running[nrow(bridge) - 1]
bridge$ymin <- c(0, head(bridge$running, -1))
bridge$ymax <- bridge$running
bridge$ymin[bridge$type %in% c("start", "end")] <- 0

# Determine colors
bridge$bar_fill <- ifelse(
    bridge$type %in% c("start", "end"),
    cols$actual,
    cols$actual_light
)
bridge$label_color <- ifelse(
    bridge$type %in% c("start", "end"),
    "white",
    ifelse(bridge$value >= 0, cols$variance_positive, cols$variance_negative)
)

# Create labels
bridge$label <- ifelse(
    bridge$type %in% c("start", "end"),
    sprintf("%.0f", bridge$ymax),
    sprintf("%+.0f", bridge$value)
)

bridge$driver <- factor(bridge$driver, levels = bridge$driver)

ggplot(bridge, aes(x = driver)) +
    geom_rect(
        aes(xmin = as.numeric(driver) - 0.4,
            xmax = as.numeric(driver) + 0.4,
            ymin = ymin, ymax = ymax),
        fill = bridge$bar_fill,
        color = cols$actual
    ) +
    geom_marquee_colored(
        data = bridge,
        mapping = aes(
            x = driver,
            y = (ymin + ymax) / 2,
            label = label,
            color = label_color
        ),
        bold = TRUE,
        size = 4
    ) +
    labs(title = "Revenue Bridge with Colored Labels", x = NULL, y = "Revenue ($M)") +
    theme_minimal()
```

## Marquee Syntax Reference

The marquee package uses a markdown-like syntax for formatting:

| Syntax | Result |
|--------|--------|
| `{.red text}` | Red colored text |
| `{#FF0000 text}` | Hex-colored text |
| `**text**` | **Bold text** |
| `*text*` | *Italic text* |
| `***text***` | ***Bold italic text*** |
| `{.red **text**}` | Bold red text |

## Installation

If marquee is not installed, install it from CRAN:

```r
install.packages("marquee")
```

## Best Practices

1. **Use sparingly** - Colored text is powerful; don't overuse it

2. **Match semantics** - Green for good, red for bad (in Western contexts)

3. **Ensure readability** - Make sure colors have sufficient contrast

4. **Consider accessibility** - Some users may have color vision deficiencies;
   use shape and position as additional indicators

5. **Be consistent** - Use the same color meanings throughout your report
