---
title: "Introduction to stwd"
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(stwd)
library(ggplot2)
library(patchwork)
```

## What is stwd?

The **stwd** package provides tools for creating presentation-ready visualizations
following [Storytelling with Data](https://www.storytellingwithdata.com/) principles.
It helps you build complete "story layouts" that combine charts with narrative text,
colored titles, and supporting context.

## Key Features

The package includes:

1. **Story Layouts** - Combine charts with titles, subtitles, narratives, and captions
2. **Story Designer** - Interactive Shiny app for designing layouts
3. **Highlight Colors** - Strategic color use (gray out noise, highlight signal)
4. **Marquee Integration** - Color-matched text in titles and labels
5. **Multi-Panel Layouts** - Connectors for showing relationships between plots

## The SWD Approach

Traditional charts put the burden on the reader to interpret meaning.
Storytelling with Data (SWD) charts guide the reader by:

- **Highlighting what matters** - Gray out the noise, color the signal
- **Embedding the legend in the title** - No separate legend box needed
- **Adding narrative context** - Explain the "so what?" alongside the chart
- **Using a clear visual hierarchy** - Big title, supporting chart, explanatory text

## Quick Start

### Create a Story Layout

```{r}
#| label: story-example
#| fig-width: 10
#| fig-height: 7
# Sample data
df <- data.frame(
    product = c("Apples", "Oranges", "Bananas", "Grapes"),
    sales = c(120, 85, 95, 110)
)

# Highlight only the winner
fill_colors <- highlight_colors(
    df$product,
    highlight = c("Apples" = "#E63946")
)

# Create the base chart
p <- ggplot(df, aes(x = reorder(product, sales), y = sales, fill = product)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = fill_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal() +
    labs(x = NULL, y = "Units Sold")

# Compose the full story
story_layout(
    plot = p,
    title = "**Q4 Sales: {#E63946 Apples} Take the Lead**",
    subtitle = "Fruit category performance, October-December 2024",
    narrative = "**Key takeaway:**

{#E63946 Apples} outsold all other fruits
this quarter, driven by seasonal demand
and promotional pricing.

**Recommendation:** Increase apple
inventory for Q1.",
    caption = "Source: Internal Sales Database"
)
```

### Use the Interactive Designer

Launch the Story Designer Shiny app to interactively design layouts:

```r
story_designer()
```

The app lets you:

- Adjust text sizes and spacing
- Preview layouts at actual export size
- Detect text clipping issues
- Generate code for your final layout

## Learn More

- [Storytelling Layouts](storytelling.html) - Complete guide to story layouts
- [Multi-Panel Layouts](layouts.html) - Connecting related charts
- [Marquee Integration](marquee.html) - Colored text labels

## References

- [Storytelling with Data](https://www.storytellingwithdata.com/)
- [Recreating the SWD Look](https://albert-rapp.de/posts/ggplot2-tips/10_recreating_swd_look/)
