---
title: "Introduction to stwd"
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(stwd)
library(ggplot2)
library(patchwork)
```

## What is stwd?

The **stwd** package provides tools for creating presentation-ready visualizations
following [Storytelling with Data](https://www.storytellingwithdata.com/) principles.
It helps you build complete "story layouts" that combine charts with narrative text,
colored titles, and supporting context.

## Key Features

The package includes:

1. **Story Designer** - Interactive Shiny app for designing layouts with real-time preview
2. **Block Functions** - Modular components (title, subtitle, narrative, caption) for patchwork composition
3. **Legend Block** - Inline colored text legends embedded in the layout
4. **Highlight Colors** - Strategic color use (gray out noise, highlight signal)
5. **Marquee Integration** - Color-matched text in titles and labels

## The STWD Approach

Traditional charts put the burden on the reader to interpret meaning.
Storytelling with Data (SWD) charts guide the reader by:

- **Highlighting what matters** - Gray out the noise, color the signal
- **Embedding the legend in the title** - No separate legend box needed
- **Adding narrative context** - Explain the "so what?" alongside the chart
- **Using a clear visual hierarchy** - Big title, supporting chart, explanatory text

## Workflow

The stwd workflow is:

1. **Create your plot** with ggplot2
2. **Design interactively** with `story_designer()`
3. **Copy the generated code** from the Code tab
4. **Paste into your Quarto document**

## Quick Start

### Step 1: Create Your Plot

```{r}
#| label: create-plot
# Sample data
df <- data.frame(
    product = c("Apples", "Oranges", "Bananas", "Grapes"),
    sales = c(120, 85, 95, 110)
)

# Highlight only the winner
fill_colors <- highlight_colors(
    df$product,
    highlight = c("Apples" = "#E63946")
)

# Create the base chart
p <- ggplot(df, aes(x = reorder(product, sales), y = sales, fill = product)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = fill_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_stwd() +
    labs(x = NULL, y = "Units Sold")
```

### Step 2: Launch the Designer

```r
story_designer(plot = p)
```

The designer lets you:

- Add and customize title, subtitle, narrative, caption
- Adjust font sizes, alignment, spacing
- Apply color palettes or manual colors
- Preview at actual export size
- Copy the generated code

### Step 3: Use the Generated Code

The designer generates patchwork code like this:

```{r}
#| label: story-example
#| fig-width: 10
#| fig-height: 7
#| fig-dpi: 150

# Block components
title_plot <- title_block(
    "**Q4 Sales: {#E63946 Apples} Take the Lead**",
    title_size = 12
)

subtitle_plot <- subtitle_block(
    "Fruit category performance, October-December 2024",
    subtitle_size = 10
)

narrative_plot <- text_narrative(
    "**Key takeaway:**

{#E63946 Apples} outsold all other fruits
this quarter, driven by seasonal demand
and promotional pricing.

**Recommendation:** Increase apple
inventory for Q1.",
    size = 10
)

caption_plot <- caption_block(
    "Source: Internal Sales Database",
    caption_size = 8
)

# Combine with patchwork
content <- p + narrative_plot + plot_layout(widths = c(0.65, 0.35))
title_plot / subtitle_plot / content / caption_plot +
    plot_layout(heights = c(0.08, 0.05, 0.82, 0.05))
```

## Story Designer Features

**Text Controls:**

- Title, subtitle, narrative, caption with marquee formatting
- Font size, alignment, line height, wrap width

**Legend Block:**

- Inline colored text legend (e.g., "HIGH | MEDIUM | LOW")
- Position above, below, left, or right of chart

**Color Palettes:**

- 9 palette packages: ggsci, MetBrewer, nord, PNWColors, rcartocolor, RColorBrewer, scico, viridis, wesanderson
- Click swatches to select specific colors
- Apply to fill, color, or both

**Manual Colors:**

- Set a default color (gray) for unassigned categories
- Assign specific colors to highlight key categories
- Perfect for the SWD approach: gray the noise, color the signal

**Validate & Export:**

- Preview at actual export dimensions
- PNG, PDF, or SVG export
- Copy generated patchwork code

## Inline Legends with legend_block()

Instead of a separate legend box, embed the legend in your layout:

```{r}
#| label: legend-example
#| fig-width: 10
#| fig-height: 6
legend_colors <- c("HIGH" = "#E63946", "MEDIUM" = "#808080", "LOW" = "#457B9D")

title_block("Performance by Category") /
  legend_block(legend_colors, halign = "right") /
  p /
  caption_block("SOURCE: Internal Data") +
  plot_layout(heights = c(0.08, 0.04, 0.83, 0.05))
```

## Learn More

- [Storytelling Layouts](storytelling.html) - Complete guide to story layouts
- [Sizing Guide](sizing.html) - Matching Quarto dimensions to Story Designer
- [Marquee Integration](marquee.html) - Colored text labels

## References

- [Storytelling with Data](https://www.storytellingwithdata.com/)
- [Recreating the SWD Look](https://albert-rapp.de/posts/ggplot2-tips/10_recreating_swd_look/)
