---
title: "Introduction to icbs"
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
```

## What is icbs?
The **icbs** package provides tools for creating visualizations that follow the
[International Business Communication Standards (IBCS)](https://www.ibcs.com/).
IBCS is a set of practical rules for designing comprehensible reports,
presentations, and dashboards.

## Key Features

The package includes:

1. **IBCS Theme & Colors** - A clean ggplot2 theme and standardized color palette
2. **Scenario Notation** - Scales for distinguishing Actual, Plan, Previous Year, and Forecast data
3. **Variance Charts** - Visualize deviations with integrated variance indicators
4. **Waterfall Charts** - Show how values build up or break down
5. **Multi-Panel Layouts** - Connectors for showing relationships between plots
6. **Marquee Integration** - Color-matched text labels

## The SUCCESS Framework

IBCS is built on the SUCCESS framework:
- **S**ay - Convey a clear message
- **U**nify - Apply consistent semantic notation
- **C**ondense - Increase information density
- **C**heck - Ensure visual integrity
- **E**xpress - Choose proper visualizations
- **S**implify - Avoid clutter
- **S**tructure - Organize content logically

## Quick Start

### Apply the IBCS Theme

```{r}
#| label: theme-example
#| fig-width: 7
#| fig-height: 4
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
    geom_boxplot(fill = ibcs_colors()$actual) +
    labs(
        title = "Fuel Efficiency by Cylinder Count",
        x = "Cylinders",
        y = "Miles per Gallon"
    ) +
    theme_ibcs()
```

### Create a Variance Chart

```{r}
#| label: variance-example
#| fig-width: 7
#| fig-height: 4
sales_data <- data.frame(
    product = c("Widget A", "Widget B", "Widget C", "Widget D"),
    actual = c(120, 85, 95, 110),
    plan = c(100, 90, 100, 100)
)

ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan,
    horizontal = TRUE
)
```

### Create a Waterfall Chart

```{r}
#| label: waterfall-example
#| fig-width: 7
#| fig-height: 4
income_data <- data.frame(
    item = c("Revenue", "COGS", "Gross Profit", "Operating Expenses", "Net Income"),
    value = c(1000, -400, NA, -350, NA),
    type = c("start", "negative", "subtotal", "negative", "end")
)

ibcs_waterfall(income_data, x = item, y = value, type = type)
```

## Learn More

- [IBCS Theme and Colors](theme-colors.html) - Detailed guide to styling
- [Scenario Notation](scenario-notation.html) - How to show AC, PL, PY, FC
- [Variance Charts](variance-charts.html) - Deviation analysis
- [Waterfall Charts](waterfall-charts.html) - Contribution analysis
- [Multi-Panel Layouts](layouts.html) - Connecting related charts
- [Marquee Integration](marquee.html) - Colored text labels

## References

- [IBCS Standards](https://www.ibcs.com/standards/)
- [IBCS SUCCESS Framework](https://www.ibcs.com/ibcs-standards-1-2/)
