---
title: "Multi-Panel Layouts with Connectors"
vignette: >
  %\VignetteIndexEntry{Multi-Panel Layouts with Connectors}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
library(patchwork)
```

## Overview

The icbs package provides functions for creating multi-panel layouts with
visual connectors that show relationships between charts. These are useful
for showing:

- How one dataset splits into subcategories
- How multiple data sources merge into a combined view
- Hierarchical or flow-based relationships

## Creating Connectors

### The make_connector() Function

The `make_connector()` function creates a standalone connector element:

```{r}
#| label: connector-split
#| fig-width: 4
#| fig-height: 4
make_connector(direction = "split", sign = "+")
```

```{r}
#| label: connector-merge
#| fig-width: 4
#| fig-height: 4
make_connector(direction = "merge", sign = "+")
```

### Connector Parameters

```{r}
#| eval: false
make_connector(
    direction = "split",   # "split" or "merge"
    sign = "+",            # Label in the center
    line_col = "black",    # Line color
    line_size = 0.5,       # Line width
    label_size = 10,       # Text size
    label_fill = "white"   # Label background
)
```

### Customizing Connectors

```{r}
#| label: custom-connector
#| fig-width: 4
#| fig-height: 4
make_connector(
    direction = "split",
    sign = "OR",
    line_col = "steelblue",
    line_size = 1,
    label_size = 8,
    label_fill = "lightblue"
)
```

## One-to-Two Split Layout

Use `one_to_split_to_two_layout()` when one chart on the left should
connect to two charts on the right (showing a breakdown or split):

```{r}
#| label: split-layout
#| fig-width: 10
#| fig-height: 6
# Create example plots
cols <- ibcs_colors()

# Total sales (left panel)
total_sales <- data.frame(
    quarter = c("Q1", "Q2", "Q3", "Q4"),
    sales = c(100, 120, 110, 130)
)

plot_total <- ggplot(total_sales, aes(x = quarter, y = sales)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Total Sales", y = "Sales ($M)") +
    theme_ibcs()

# Online sales (top right)
online_sales <- data.frame(
    quarter = c("Q1", "Q2", "Q3", "Q4"),
    sales = c(60, 80, 75, 90)
)

plot_online <- ggplot(online_sales, aes(x = quarter, y = sales)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Online Sales", y = "Sales ($M)") +
    theme_ibcs()

# Retail sales (bottom right)
retail_sales <- data.frame(
    quarter = c("Q1", "Q2", "Q3", "Q4"),
    sales = c(40, 40, 35, 40)
)

plot_retail <- ggplot(retail_sales, aes(x = quarter, y = sales)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Retail Sales", y = "Sales ($M)") +
    theme_ibcs()

# Combine with split layout
one_to_split_to_two_layout(
    list(plot_total, plot_online, plot_retail),
    first_level_sign = "="
)
```

## Two-to-One Merge Layout

Use `two_to_merge_to_one_layout()` when two charts on the left should
connect to one chart on the right (showing aggregation):

```{r}
#| label: merge-layout
#| fig-width: 10
#| fig-height: 6
# Product A (top left)
product_a <- data.frame(
    month = c("Jan", "Feb", "Mar"),
    revenue = c(50, 55, 60)
)

plot_a <- ggplot(product_a, aes(x = month, y = revenue)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Product A", y = "Revenue ($K)") +
    theme_ibcs()

# Product B (bottom left)
product_b <- data.frame(
    month = c("Jan", "Feb", "Mar"),
    revenue = c(30, 35, 40)
)

plot_b <- ggplot(product_b, aes(x = month, y = revenue)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Product B", y = "Revenue ($K)") +
    theme_ibcs()

# Combined (right)
combined <- data.frame(
    month = c("Jan", "Feb", "Mar"),
    revenue = c(80, 90, 100)
)

plot_combined <- ggplot(combined, aes(x = month, y = revenue)) +
    geom_col(fill = cols$actual, width = 0.6) +
    labs(title = "Total Revenue", y = "Revenue ($K)") +
    theme_ibcs()

# Combine with merge layout
two_to_merge_to_one_layout(
    list(plot_a, plot_b, plot_combined),
    first_level_sign = "+"
)
```

## Adding Panel Labels

Use `add_panel_labels()` to add letter labels (A, B, C) to multi-panel
figures:

```{r}
#| label: panel-labels
#| fig-width: 8
#| fig-height: 4
# Create a simple multi-panel plot
p1 <- ggplot(mtcars, aes(x = mpg)) +
    geom_histogram(fill = cols$actual, bins = 10) +
    labs(title = "MPG Distribution") +
    theme_ibcs()

p2 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(color = cols$actual) +
    labs(title = "Weight vs MPG") +
    theme_ibcs()

combined_plot <- p1 + p2

# Add labels
add_panel_labels(combined_plot)
```

### Label Styles

```{r}
#| label: label-styles
#| fig-width: 8
#| fig-height: 4
# Lowercase labels
add_panel_labels(combined_plot, style = "lower")
```

```{r}
#| label: custom-labels
#| fig-width: 8
#| fig-height: 4
# Custom labels
add_panel_labels(combined_plot, labels = c("Fig 1", "Fig 2"))
```

### Label Customization

```{r}
#| label: label-position
#| fig-width: 8
#| fig-height: 4
add_panel_labels(
    combined_plot,
    style = "upper",
    fontface = "bold.italic",
    size = 16,
    x = 0.05,
    y = 0.95
)
```

## Building Custom Layouts

You can combine connectors with patchwork for custom layouts:

```{r}
#| label: custom-layout
#| fig-width: 12
#| fig-height: 6
# Create three related plots
revenue_plot <- ggplot(data.frame(x = 1:4, y = c(100, 120, 115, 130)),
                        aes(x = x, y = y)) +
    geom_col(fill = cols$actual) +
    labs(title = "Revenue") +
    theme_ibcs()

cost_plot <- ggplot(data.frame(x = 1:4, y = c(60, 70, 65, 75)),
                     aes(x = x, y = y)) +
    geom_col(fill = cols$actual) +
    labs(title = "Costs") +
    theme_ibcs()

profit_plot <- ggplot(data.frame(x = 1:4, y = c(40, 50, 50, 55)),
                       aes(x = x, y = y)) +
    geom_col(fill = cols$actual) +
    labs(title = "Profit") +
    theme_ibcs()

# Custom layout with connector
layout <- "
AACBB
AADBB
"

connector1 <- make_connector(direction = "split", sign = "-")

revenue_plot + profit_plot + connector1 + cost_plot +
    plot_layout(design = layout)
```

## Practical Example: Sales Breakdown

```{r}
#| label: practical-example
#| fig-width: 11
#| fig-height: 7
# Regional breakdown example
total_by_year <- data.frame(
    year = 2020:2024,
    sales = c(800, 850, 900, 950, 1000)
)

north_by_year <- data.frame(
    year = 2020:2024,
    sales = c(300, 320, 340, 360, 380)
)

south_by_year <- data.frame(
    year = 2020:2024,
    sales = c(500, 530, 560, 590, 620)
)

p_total <- ggplot(total_by_year, aes(x = year, y = sales)) +
    geom_line(color = cols$actual, linewidth = 1) +
    geom_point(color = cols$actual, size = 3) +
    labs(title = "Total Company Sales", y = "Sales ($M)", x = NULL) +
    scale_y_continuous(limits = c(0, 1100)) +
    theme_ibcs()

p_north <- ggplot(north_by_year, aes(x = year, y = sales)) +
    geom_line(color = cols$actual, linewidth = 1) +
    geom_point(color = cols$actual, size = 3) +
    labs(title = "North Region", y = "Sales ($M)", x = NULL) +
    scale_y_continuous(limits = c(0, 700)) +
    theme_ibcs()

p_south <- ggplot(south_by_year, aes(x = year, y = sales)) +
    geom_line(color = cols$actual, linewidth = 1) +
    geom_point(color = cols$actual, size = 3) +
    labs(title = "South Region", y = "Sales ($M)", x = NULL) +
    scale_y_continuous(limits = c(0, 700)) +
    theme_ibcs()

one_to_split_to_two_layout(
    list(p_total, p_north, p_south),
    first_level_sign = "="
)
```

## Best Practices

1. **Use connectors sparingly** - Only when showing explicit relationships

2. **Match scales when possible** - Helps viewers compare across panels

3. **Choose meaningful signs** - Use "+" for sums, "=" for breakdowns,
   "×" for products, etc.

4. **Label panels** - Use `add_panel_labels()` for reference in text

5. **Keep layouts simple** - Complex layouts can confuse rather than clarify
