---
title: "IBCS Scenario Notation"
vignette: >
  %\VignetteIndexEntry{IBCS Scenario Notation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
```

## Understanding Scenario Notation

IBCS defines a standardized visual language for distinguishing different types
of business data scenarios. The notation uses fill patterns and borders to
instantly communicate what kind of data is being shown.

### The Four Scenarios

| Scenario | Abbreviation | Visual Style | Description |
|----------|--------------|--------------|-------------|
| Actual | AC | Solid dark fill | Measured data from current period |
| Plan/Budget | PL | Outlined (hollow) | Target or budgeted values |
| Previous Year | PY | Lighter solid fill | Historical comparison data |
| Forecast | FC | Hatched pattern* | Projected future values |
\* In this package, forecast is approximated with a lighter fill and border

```{r}
#| label: notation-reference
#| fig-width: 8
#| fig-height: 3
# Visual reference for scenario notation
scenarios <- data.frame(
    scenario = factor(c("AC\n(Actual)", "PL\n(Plan)", "PY\n(Previous)", "FC\n(Forecast)"),
                      levels = c("AC\n(Actual)", "PL\n(Plan)", "PY\n(Previous)", "FC\n(Forecast)")),
    value = c(100, 100, 100, 100)
)

cols <- ibcs_colors()

ggplot(scenarios, aes(x = scenario, y = value)) +
    geom_col(
        fill = c(cols$actual, NA, cols$previous, cols$actual_light),
        color = cols$actual,
        linewidth = c(0.3, 1.5, 0.3, 0.5),
        width = 0.6
    ) +
    labs(title = "IBCS Scenario Notation Reference", x = NULL, y = NULL) +
    theme_ibcs() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

## Using Scenario Scales

### scale_fill_ibcs_scenario()

Use this scale when mapping scenario types to fill aesthetics:
```{r}
#| label: fill-scale
#| fig-width: 7
#| fig-height: 4
quarterly <- data.frame(
    quarter = rep(c("Q1", "Q2", "Q3", "Q4"), 2),
    scenario = rep(c("AC", "PY"), each = 4),
    revenue = c(120, 135, 125, 140, 110, 120, 115, 130)
)

ggplot(quarterly, aes(x = quarter, y = revenue, fill = scenario)) +
    geom_col(position = "dodge", width = 0.7, color = ibcs_colors()$actual) +
    scale_fill_ibcs_scenario() +
    labs(
        title = "Quarterly Revenue: Actual vs Previous Year",
        x = NULL, y = "Revenue ($M)"
    ) +
    theme_ibcs()
```

### scale_color_ibcs_scenario()

Use for border colors, especially important for Plan (outlined bars):

```{r}
#| label: color-scale
#| fig-width: 7
#| fig-height: 4
monthly <- data.frame(
    month = rep(c("Jan", "Feb", "Mar"), 2),
    scenario = rep(c("AC", "PL"), each = 3),
    sales = c(100, 110, 105, 95, 100, 100)
)

ggplot(monthly, aes(x = month, y = sales, fill = scenario, color = scenario)) +
    geom_col(position = "dodge", width = 0.7, linewidth = 1) +
    scale_fill_ibcs_scenario() +
    scale_color_ibcs_scenario() +
    labs(
        title = "Monthly Sales: Actual vs Plan",
        x = NULL, y = "Sales (units)"
    ) +
    theme_ibcs()
```

## The ibcs_bar_chart() Function
For convenience, use `ibcs_bar_chart()` which handles all the notation automatically:

```{r}
#| label: bar-chart
#| fig-width: 8
#| fig-height: 4
performance <- data.frame(
    region = rep(c("North", "South", "East", "West"), each = 2),
    scenario = rep(c("AC", "PL"), 4),
    value = c(150, 140, 120, 130, 180, 170, 95, 100)
)

ibcs_bar_chart(performance, x = region, y = value, scenario = scenario)
```

## Variance Visualization

When showing variances (differences between scenarios), use the dedicated
variance colors:

### scale_fill_ibcs_variance()

```{r}
#| label: variance-scale
#| fig-width: 7
#| fig-height: 4
variance_data <- data.frame(
    category = c("Product A", "Product B", "Product C", "Product D"),
    variance = c(15, -10, 5, -3),
    direction = c("positive", "negative", "positive", "negative")
)

ggplot(variance_data, aes(x = reorder(category, variance), y = variance, fill = direction)) +
    geom_col(width = 0.6) +
    scale_fill_ibcs_variance() +
    coord_flip() +
    labs(
        title = "Variance to Plan by Product",
        x = NULL, y = "Variance (%)"
    ) +
    theme_ibcs() +
    theme(legend.position = "none")
```

## Multi-Scenario Comparisons

### Actual vs Plan vs Previous Year

```{r}
#| label: multi-scenario
#| fig-width: 9
#| fig-height: 5
multi <- data.frame(
    month = rep(c("Jan", "Feb", "Mar", "Apr"), 3),
    scenario = rep(c("AC", "PL", "PY"), each = 4),
    value = c(
        100, 110, 115, 120,  # Actual
        95, 105, 110, 115,   # Plan
        90, 95, 100, 105     # Previous Year
    )
)

ibcs_bar_chart(multi, x = month, y = value, scenario = scenario) +
    labs(title = "Monthly Performance: Three Scenario Comparison")
```

## Time Series with Forecast

When showing time series that transition from actual to forecast:

```{r}
#| label: time-series
#| fig-width: 9
#| fig-height: 4
cols <- ibcs_colors()

timeline <- data.frame(
    month = factor(month.abb, levels = month.abb),
    value = c(100, 105, 98, 110, 115, 108, 120, 125, 130, 128, 135, 140),
    scenario = c(rep("AC", 6), rep("FC", 6))
)

ggplot(timeline, aes(x = month, y = value, fill = scenario, color = scenario)) +
    geom_col(width = 0.7, linewidth = 0.5) +
    geom_vline(xintercept = 6.5, linetype = "dashed", color = cols$previous) +
    annotate("text", x = 3.5, y = 145, label = "Actual", color = cols$actual) +
    annotate("text", x = 9.5, y = 145, label = "Forecast", color = cols$actual_light) +
    scale_fill_ibcs_scenario() +
    scale_color_ibcs_scenario() +
    labs(
        title = "Annual Sales Projection",
        x = NULL, y = "Sales ($K)"
    ) +
    theme_ibcs() +
    theme(legend.position = "none")
```

## Best Practices

### Consistent Application

Always use the same notation across all visualizations in a report:

- AC (Actual) = solid dark
- PL (Plan) = outlined
- PY (Previous) = lighter solid
- FC (Forecast) = hatched/lighter with border

### Integrated Legends

IBCS recommends integrating legends into the visualization rather than using
separate legend boxes. Label scenarios directly on or near the data:

```{r}
#| label: integrated-legend
#| fig-width: 8
#| fig-height: 4
cols <- ibcs_colors()

comparison <- data.frame(
    period = c("2023", "2024"),
    actual = c(100, 120),
    plan = c(95, 115)
)

ggplot(comparison) +
    # Plan bars (outlined)
    geom_col(
        aes(x = period, y = plan),
        fill = NA, color = cols$plan, linewidth = 1.2, width = 0.5
    ) +
    # Actual bars (solid, narrower, overlaid)
    geom_col(
        aes(x = period, y = actual),
        fill = cols$actual, width = 0.3
    ) +
    # Integrated labels
    annotate("text", x = 2.3, y = 120, label = "AC", hjust = 0, fontface = "bold", size = 3) +
    annotate("text", x = 2.3, y = 115, label = "PL", hjust = 0, size = 3) +
    labs(title = "Year-over-Year Comparison", x = NULL, y = "Revenue ($M)") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_ibcs()
```
