---
title: "Debugging Layouts with story_sitrep()"
---

```{r}
#| include: false
devtools::load_all()
library(ggplot2)
library(patchwork)
```

## Introduction

When building complex story layouts with `story_layout()`, it can be difficult to understand how the various components fit together. The `story_sitrep()` function provides a "situation report" that shows you exactly what's in your layout before you render it.

This is useful for:

- **Debugging** - Understanding why elements overlap or look wrong
- **Planning** - Previewing layout proportions before committing
- **Documentation** - Seeing all parameters at a glance

## Basic Usage

`story_sitrep()` takes the same arguments as `story_layout()` and returns a diagnostic object:

```{r}
sitrep <- story_sitrep(
    title = "My Report Title",
    subtitle = "Supporting context here",
    narrative = "Key insights and analysis",
    caption = "Source: Data Team"
)

sitrep
```

The output shows:

1. **Components** - Which elements are included, their type (text or ggplot), and size in points
2. **Layout Proportions** - How much vertical space each element takes
3. **Layout Diagram** - ASCII visualization of the structure
4. **Tips** - SWD-inspired suggestions for improvement

## Understanding the Layout Diagram

The ASCII diagram shows how your layout will be structured. When you have both a plot and narrative, the diagram shows their arrangement:

```{r}
# Create a placeholder plot for demonstration
p <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
    geom_col() +
    theme_minimal()

# Layout with narrative on the RIGHT
story_sitrep(
    plot = p,
    title = "Sales Analysis",
    narrative = "Commentary here",
    narrative_position = "right"
)
```

Compare with narrative on the LEFT - notice how NARRATIVE and PLOT swap positions:

```{r}
story_sitrep(
    plot = p,
    title = "Sales Analysis",
    narrative = "Commentary here",
    narrative_position = "left"
)
```

Or narrative at the BOTTOM (stacked vertically):

```{r}
story_sitrep(
    plot = p,
    title = "Sales Analysis",
    narrative = "Commentary here",
    narrative_position = "bottom",
    narrative_width = 0.2
)
```

## Checking Proportions

The sitrep helps you verify that your height allocations make sense:

```{r}
# Default proportions
story_sitrep(
    title = "Title",
    subtitle = "Subtitle",
    caption = "Caption"
)
```

If you have a long title that wraps to multiple lines, you may need more height:

```{r}
# Increased title height for wrapping text
story_sitrep(
    title = "This is a Very Long Title That Will Wrap to Multiple Lines",
    subtitle = "Subtitle",
    title_height = 0.20
)
```

## SWD-Inspired Tips

The sitrep provides suggestions based on Storytelling with Data principles:

```{r}
# Small title - sitrep suggests larger for visual hierarchy
story_sitrep(
    title = "Small Title",
    title_size = 10
)
```

```{r}
# Plot without narrative - sitrep suggests adding context
story_sitrep(
    plot = p,
    title = "Report"
)
```

```{r}
# Plot without source - sitrep suggests adding credibility
story_sitrep(
    plot = p,
    title = "Report",
    narrative = "Analysis"
)
```

## Workflow: Sitrep Then Build

A recommended workflow is to check your sitrep before rendering:

```{r}
#| fig-width: 10
#| fig-height: 7

# Step 1: Check the sitrep
my_sitrep <- story_sitrep(
    title = "**Q4 Results: {#E63946 Sales} Exceed Target**",
    subtitle = "Regional performance summary",
    narrative = "Key findings go here",
    narrative_position = "right",
    narrative_width = 0.35,
    caption = "Source: Finance"
)

my_sitrep
```

```{r}
#| fig-width: 10
#| fig-height: 7

# Step 2: If proportions look good, build the actual layout
df <- data.frame(
    region = c("North", "South", "East", "West"),
    sales = c(120, 95, 110, 85)
)

p <- ggplot(df, aes(x = reorder(region, sales), y = sales)) +
    geom_col(fill = "#E63946") +
    coord_flip() +
    theme_minimal() +
    labs(x = NULL, y = "Sales ($M)")

story_layout(
    plot = p,
    title = "**Q4 Results: {#E63946 Sales} Exceed Target**",
    subtitle = "Regional performance summary",
    narrative = "**North region** led all markets
with $120M in sales.

West underperformed at $85M,
requiring attention in Q1.",
    narrative_position = "right",
    narrative_width = 0.35,
    caption = "Source: Finance"
)
```

## Accessing Sitrep Data Programmatically

The sitrep is an S7 object. You can access its properties directly:

```{r}
sitrep <- story_sitrep(
    title = "Report",
    subtitle = "Details",
    narrative = "Analysis"
)

# Access the components table
sitrep@components

# Access layout proportions
sitrep@layout$content_height

# Check for tips
sitrep@tips
```

## Preventing Text Clipping with auto_heights

A common issue is titles or subtitles getting clipped when they wrap to multiple lines. The `auto_heights` parameter (default `TRUE`) automatically calculates appropriate heights:

```{r}
#| fig-width: 12
#| fig-height: 9

# Long title that wraps - auto_heights handles it automatically
story_layout(
    plot = p,
    title = "**{#E63946 10 warehouses} need immediate attention for error rates**",
    subtitle = "What are the next steps at our highest volume sites?",
    caption = "Source: Operations Dashboard",
    auto_heights = TRUE,  # Default - calculates heights from text length
    output_width = 12     # Match your ggsave width for accurate estimates
)
```

For manual control, set `auto_heights = FALSE` and specify heights explicitly:

```{r}
#| eval: false

story_layout(
    plot = p,
    title = "Short title",
    auto_heights = FALSE,
    title_height = 0.10,
    subtitle_height = 0.06
)
```

**Best practice:** Keep titles under 60 characters. Concise, punchy titles are both easier to read AND less likely to have layout issues.

## Summary

| Function | Purpose |
|----------|---------|
| `story_sitrep()` | Preview layout structure before rendering |
| `@components` | Data frame of included elements |
| `@layout` | List of proportions and positions |
| `@diagram` | ASCII visualization |
| `@tips` | SWD-inspired suggestions |

Use `story_sitrep()` to debug layouts, verify proportions, and get SWD-inspired suggestions before rendering.
