---
title: "Storytelling with Data Layouts"
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
devtools::load_all()
library(ggplot2)
library(marquee)
library(patchwork)
```

## Overview

The stwd package includes functions inspired by
[Storytelling with Data](https://www.storytellingwithdata.com/) principles
for creating presentation-ready charts. These tools help you:
- Combine charts with narrative text
- Create self-documenting titles with colored category names
- Use color strategically for emphasis
- Build complete "story" layouts with one function call

## The SWD Approach

Traditional charts put the burden on the reader to interpret meaning.
SWD-style charts guide the reader by:

1. **Highlighting what matters** - Gray out the noise, color the signal
2. **Embedding the legend in the title** - No separate legend box needed
3. **Adding narrative context** - Explain the "so what?" alongside the chart
4. **Using a clear visual hierarchy** - Big title, supporting chart, explanatory text

## Quick Start: story_layout()
The `story_layout()` function combines all elements into a complete
presentation-ready visualization. By default, titles are large and prominent
(16pt), following the [Storytelling with Data](https://albert-rapp.de/posts/ggplot2-tips/10_recreating_swd_look/10_recreating_swd_look.html) approach.

```{r}
#| label: quick-start
#| fig-width: 10
#| fig-height: 7
df <- data.frame(
    product = c("Apples", "Oranges", "Bananas", "Grapes"),
    sales = c(120, 85, 95, 110)
)

# Highlight only the winner
fill_colors <- highlight_colors(
    df$product,
    highlight = c("Apples" = "#E63946")
)

# Create the base chart
p <- ggplot(df, aes(x = reorder(product, sales), y = sales, fill = product)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_manual(values = fill_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal() +
    labs(x = NULL, y = "Units Sold")

# Compose the full story - title is large by default (16pt)
story_layout(
    plot = p,
    title = "**Q4 Sales: {#E63946 Apples} Take the Lead**",
    subtitle = "Fruit category performance, October-December 2024",
    narrative = "**Key takeaway:**

{#E63946 Apples} outsold all other fruits
this quarter, driven by seasonal demand
and promotional pricing.

Grapes came in second at 110 units,
while Oranges lagged at just 85.

**Recommendation:** Increase apple
inventory for Q1 to meet expected demand.",
    caption = "Source: Internal Sales Database"
)
```

## Building Blocks

### highlight_colors(): Strategic Color Use

Instead of coloring everything, highlight only what matters:

```{r}
#| label: highlight-colors
categories <- c("A", "B", "C", "D", "E")

# Single highlight with default blue
highlight_colors(categories, "C")
```

```{r}
#| label: highlight-multi
# Multiple highlights with custom colors
highlight_colors(
    categories,
    highlight = c("A" = "#E63946", "D" = "#2A9D8F")
)
```

### marquee_title(): Self-Documenting Titles

Embed colored category names directly in your title:

```{r}
#| label: marquee-title
fill_colors <- c(
    "Revenue" = "#2A9D8F",
    "Costs" = "#E63946",
    "Profit" = "#1E90FF"
)

marquee_title(
    "{Revenue} grew faster than {Costs}, boosting {Profit}",
    fill_colors
)
```

Use with `theme_marquee()` to render the colors:

```{r}
#| label: title-example
#| fig-width: 8
#| fig-height: 5
df <- data.frame(
    metric = c("Revenue", "Costs", "Profit"),
    value = c(150, 100, 50)
)

fill_colors <- c(
    "Revenue" = "#2A9D8F",
    "Costs" = "#E76F51",
    "Profit" = "#264653"
)

ggplot(df, aes(x = metric, y = value, fill = metric)) +
    geom_col(width = 0.6) +
    scale_fill_manual(values = fill_colors, guide = "none") +
    labs(
        title = marquee_title(
            "{Revenue} minus {Costs} equals {Profit}",
            fill_colors
        ),
        x = NULL, y = "$ Millions"
    ) +
    theme_minimal() +
    theme_marquee()
```

### text_narrative(): Standalone Text Blocks

Create text panels for use with patchwork:

```{r}
#| label: text-narrative
#| fig-width: 4
#| fig-height: 3
text_narrative(
    text = "**Key insight:**

This quarter showed {#E63946 strong growth}
in our core markets.

Action items:
- Expand capacity
- Hire additional staff",
    size = 3.5
)
```

### Modular Block Functions

Each component of a story layout is available as a standalone function,
giving you full control over sizing, positioning, and styling.

#### title_block(): Large Prominent Titles

```{r}
#| label: title-block
#| fig-width: 10
#| fig-height: 1
title_block(
    "**Quarterly Performance Review**",
    title_size = 18
)
```

#### subtitle_block(): Supporting Context

```{r}
#| label: subtitle-block
#| fig-width: 10
#| fig-height: 0.6
subtitle_block(
    "{#2A9D8F Revenue} up 15% | {#E63946 Costs} down 3%",
    subtitle_size = 12
)
```

#### caption_block(): Source Attribution

```{r}
#| label: caption-block
#| fig-width: 10
#| fig-height: 0.5
caption_block("Source: Finance Department | Q4 2024")
```

#### Combining Blocks Manually

For maximum control, combine blocks using patchwork:

```{r}
#| label: manual-blocks
#| fig-width: 10
#| fig-height: 6
df <- data.frame(x = c("A", "B", "C"), y = c(30, 50, 20))

chart <- ggplot(df, aes(x, y, fill = x)) +
    geom_col() +
    scale_fill_manual(values = c(A = "#E63946", B = "#CCCCCC", C = "#CCCCCC"),
                      guide = "none") +
    theme_minimal()

# Stack components with custom heights
title_block("**{#E63946 Category A} Leads the Pack**", title_size = 18) /
subtitle_block("Year-over-year comparison by category") /
chart /
caption_block("Source: Internal Analytics") +
    plot_layout(heights = c(0.08, 0.05, 0.83, 0.04))


```

## Manual Composition with Patchwork

For full control, build layouts manually:

```{r}
#| label: manual-patchwork
#| fig-width: 11
#| fig-height: 6
# Data
df <- data.frame(
    quarter = c("Q1", "Q2", "Q3", "Q4"),
    sales = c(85, 92, 88, 120)
)

# Colors - highlight Q4
fill_colors <- highlight_colors(
    df$quarter,
    highlight = c("Q4" = "#E63946")
)

# Main chart
chart <- ggplot(df, aes(x = quarter, y = sales, fill = quarter)) +
    geom_col(width = 0.7) +
    geom_hline(yintercept = 100, linetype = "dashed", color = "#666666") +
    annotate("text", x = 4.5, y = 102, label = "Target",
             hjust = 1, size = 3, color = "#666666") +
    scale_fill_manual(values = fill_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal() +
    labs(x = NULL, y = "Sales (units)")

# Narrative panel
narrative <- text_narrative(
    text = "**The {#E63946 Q4} breakthrough**

After three quarters of hovering
around 85-92 units, {#E63946 Q4}
finally broke through the 100-unit
target with **120 units sold**.

This 30% jump was driven by:

- Holiday promotions
- New product launch
- Expanded distribution

**Next steps:** Maintain momentum
into Q1 with continued marketing.",
    size = 3.2,
    valign = "top"
)

# Compose with patchwork
(chart + narrative) +
    plot_layout(widths = c(0.6, 0.4)) +
    plot_annotation(
        title = "**Annual Sales Trend: {#E63946 Q4} Exceeds Target**",
        subtitle = "Quarterly unit sales with 100-unit target line",
        caption = "Source: Sales Database | Analysis: Product Team",
        theme = theme(
            plot.title = element_marquee(size = 14),
            plot.subtitle = element_marquee(size = 10),
            plot.caption = element_marquee(size = 8)
        )
    )
```

## Colored Legend Text

Match legend text colors to bar fills using `scale_fill_marquee()`:

```{r}
#| label: colored-legend
#| fig-width: 8
#| fig-height: 5
df <- data.frame(
    category = c("Electronics", "Clothing", "Food", "Home"),
    sales = c(450, 320, 280, 190)
)

fill_colors <- c(
    "Electronics" = "#264653",
    "Clothing" = "#2A9D8F",
    "Food" = "#E9C46A",
    "Home" = "#E76F51"
)

ggplot(df, aes(x = reorder(category, sales), y = sales, fill = category)) +
    geom_col(width = 0.7) +
    coord_flip() +
    scale_fill_marquee(fill_colors, bold = TRUE) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal() +
    theme(legend.text = element_marquee()) +
    labs(
        title = "Sales by Category",
        x = NULL, y = "Revenue ($K)", fill = NULL
    )
```

## Adding Colored Labels to Bars

Use `marquee_labels_from_colors()` to create bar labels that match fills:

```{r}
#| label: bar-labels
#| fig-width: 8
#| fig-height: 5
df <- data.frame(
    category = c("North", "South", "East", "West"),
    value = c(85, 120, 95, 110)
)

fill_colors <- c(
    "North" = "#457B9D",
    "South" = "#E63946",
    "East" = "#2A9D8F",
    "West" = "#F4A261"
)

# Add formatted labels
df$bar_label <- marquee_labels_from_colors(df$category, fill_colors, bold = TRUE)

ggplot(df, aes(x = category, y = value, fill = category)) +
    geom_col(width = 0.7) +
    geom_marquee(aes(label = bar_label, y = 5), size = 4) +
    scale_fill_manual(values = fill_colors, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal() +
    labs(title = "Regional Performance", x = NULL, y = "Score")
```

## Complete Example: Executive Dashboard Panel

Putting it all together for a board-ready visual:

```{r}
#| label: executive-example
#| fig-width: 12
#| fig-height: 7
# Performance data
performance <- data.frame(
    region = c("North America", "Europe", "Asia Pacific", "Latin America"),
    actual = c(245, 180, 210, 85),
    target = c(220, 190, 200, 100)
)

performance$variance <- performance$actual - performance$target
performance$var_pct <- round(100 * performance$variance / performance$target, 1)

# Highlight regions that beat target
highlights <- setNames(
    ifelse(performance$variance >= 0, "#2A9D8F", "#CCCCCC"),
    performance$region
)
highlights["Asia Pacific"] <- "#1E90FF"  # Extra highlight for story

# Chart
chart <- ggplot(
    performance,
    aes(x = reorder(region, actual), y = actual, fill = region)
) +
    geom_col(width = 0.7) +
    geom_point(aes(y = target), shape = 4, size = 4, stroke = 1.5) +
    coord_flip() +
    scale_fill_manual(values = highlights, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_minimal() +
    labs(x = NULL, y = "Revenue ($M)")

# Compose
story_layout(
    plot = chart,
    title = "**Regional Revenue: {#2A9D8F North America} and {#1E90FF Asia Pacific} Exceed Targets**",
    subtitle = "FY2024 actual revenue vs. targets (X marks target)",
    narrative = "**Executive Summary**

Two of four regions exceeded their
annual revenue targets:

{#2A9D8F **North America**}: +$25M (+11%)
Driven by enterprise software deals

{#1E90FF **Asia Pacific**}: +$10M (+5%)
Strong growth in cloud services

**Underperformers:**

Europe missed by $10M due to
currency headwinds.

Latin America fell short by $15M;
restructuring plan underway.

**FY2025 Outlook:** Cautiously
optimistic with 8% growth target.",
    narrative_width = 0.38,
    caption = "Source: Finance | Prepared for Q4 Board Meeting"
)
```

## Tips for Effective Storytelling Charts

1. **Lead with the insight** - Your title should state the conclusion, not just describe the data

2. **Use color sparingly** - Gray is your friend; color only what matters

3. **Integrate the legend** - Embed category colors in titles or narrative text

4. **Provide context** - The narrative panel explains "so what?" and "now what?"

5. **Design for your audience** - Executives want conclusions; analysts want details

6. **Keep text readable** - Don't shrink text to fit more in; simplify instead

## Function Reference

| Function | Purpose |
|----------|---------|
| `story_layout()` | One-call complete story composition |
| `title_block()` | Create large, prominent title plots (16pt default) |
| `subtitle_block()` | Create subtitle plots with context |
| `caption_block()` | Create caption/source plots |
| `text_narrative()` | Create standalone text blocks |
| `highlight_colors()` | Gray out non-essential categories |
| `marquee_title()` | Embed colored names in titles |
| `scale_fill_marquee()` | Match legend text to fill colors |
| `marquee_labels_from_colors()` | Create colored bar labels |
| `theme_marquee()` | Enable marquee for all text elements |
