---
title: "Storytelling with Data Layouts"
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
devtools::load_all()
library(ggplot2)
library(marquee)
library(patchwork)
```

## Overview

The stwd package provides block functions for building presentation-ready story layouts. Use `story_designer()` to interactively design layouts, then copy the generated patchwork code.

## Block Functions

### title_block()

Creates a large, prominent title. Supports marquee syntax for colored text.

```{r}
#| label: title-block
#| fig-width: 10
#| fig-height: 1
title_block(
    "**Q4 Sales: {#E63946 Apples} Take the Lead**",
    title_size = 10
)
```

### subtitle_block()

Adds supporting context below the title.

```{r}
#| label: subtitle-block
#| fig-width: 10
#| fig-height: 1
subtitle_block(
    "{#2A9D8F Revenue} up 15% | {#E63946 Costs} down 3%",
    subtitle_size = 10
)
```

### text_narrative()

Creates a text panel for the "so what?" explanation.

```{r}
#| label: text-narrative
#| fig-width: 4
#| fig-height: 1.5

text_narrative(
    text = "**Key insight:**
        {#E63946 Strong growth} in core markets.",
    size = 10
)
```

### caption_block()

Adds source attribution at the bottom.

```{r}
#| label: caption-block
#| fig-width: 10
#| fig-height: .8
caption_block("Source: Finance Department | Q4 2024", caption_size = 6)
```

### legend_block()

Creates an inline colored text legend instead of a traditional legend box.

```{r}
#| label: legend-block
#| fig-width: 10
#| fig-height: .8
legend_colors <- c("HIGH" = "#E63946", "MEDIUM" = "#808080", "LOW" = "#457B9D")

legend_block(legend_colors, halign = "right", sep = " | ")
```

## Combining Blocks with Patchwork

Stack blocks vertically with `/` and set proportional heights:

```{r}
#| label: combined-example
#| fig-width: 8
#| fig-height: 9    
df <- data.frame(x = c("A", "B", "C"), y = c(30, 50, 20))

chart <- ggplot(df, aes(x, y, fill = x)) +
    geom_col() +
    scale_fill_manual(values = c(A = "#E63946", B = "#CCCCCC", C = "#CCCCCC"),
                      guide = "none") +
    theme_minimal() +
    labs(x = NULL, y = NULL)

title_block("**{#E63946 Category A} Leads**", title_size = 12) /
subtitle_block("Year-over-year comparison", subtitle_size = 10) /
chart /
caption_block("Source: Internal Analytics") +
    plot_layout(heights = c(0.08, 0.06, 0.78, 0.08))
```

## Adding a Narrative Panel

Place chart and narrative side by side with `+` and set proportional widths:

```{r}
#| label: with-narrative
#| fig-width: 8
#| fig-height: 9
narrative <- text_narrative(
    "**Summary:**

{#E63946 Category A} leads with 50 units.",
    size = 8
)

content <- chart + narrative + plot_layout(widths = c(0.65, 0.35))

title_block("**{#E63946 Category A} Leads**", title_size = 12) /
subtitle_block("Year-over-year comparison", subtitle_size = 10) /
content /
caption_block("Source: Internal Analytics") +
    plot_layout(heights = c(0.08, 0.06, 0.78, 0.08))
```

## Helper Functions

| Function | Purpose |
|----------|---------|
| `highlight_colors()` | Gray out non-essential categories |
| `theme_stwd()` | Clean SWD theme (horizontal grid only) |

## Workflow

1. Create your ggplot chart
2. Launch `story_designer(plot = p)`
3. Adjust text, sizes, and layout in the app
4. Copy the generated code from the Code tab
5. Paste into your Quarto document with matching `fig-width`, `fig-height`, `fig-dpi`

See the [Sizing Guide](sizing.html) for details on matching dimensions.
