---
title: "IBCS Theme and Colors"
vignette: >
  %\VignetteIndexEntry{IBCS Theme and Colors}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
```
## The IBCS Color Palette

IBCS recommends a minimal, purposeful use of color. The standard palette reserves
bright colors (red and green) exclusively for variances, using neutral grays for
all other data representation.

### Accessing Colors

Use `ibcs_colors()` to get the standard palette:

```{r}
#| label: colors-list
cols <- ibcs_colors()
str(cols)
```

### Color Meanings

| Color | Name | Usage |
|-------|------|-------|
| `r cols$actual` | `actual` | Current period measured data |
| `r cols$actual_light` | `actual_light` | Forecast data (hatched approximation) |
| `r cols$previous` | `previous` | Previous period data |
| `r cols$plan` | `plan` | Planned/budgeted values |
| `r cols$variance_positive` | `variance_positive` | Favorable variances only |
| `r cols$variance_negative` | `variance_negative` | Unfavorable variances only |
| `r cols$highlight` | `highlight` | Special emphasis |

### Visual Color Reference

```{r}
#| label: color-swatches
#| fig-width: 8
#| fig-height: 3
cols <- ibcs_colors()

color_df <- data.frame(
    name = c("actual", "actual_light", "previous", "plan",
             "variance_positive", "variance_negative", "highlight"),
    color = c(cols$actual, cols$actual_light, cols$previous, cols$plan,
              cols$variance_positive, cols$variance_negative, cols$highlight),
    y = 1
)
color_df$name <- factor(color_df$name, levels = color_df$name)

ggplot(color_df, aes(x = name, y = y, fill = color)) +
    geom_tile(width = 0.9, height = 0.8) +
    geom_text(aes(label = color), color = "white", fontface = "bold", size = 3) +
    scale_fill_identity() +
    theme_void() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        plot.margin = margin(10, 10, 10, 10)
    ) +
    labs(title = "IBCS Color Palette")
```

## The IBCS Theme

`theme_ibcs()` provides a clean, minimal ggplot2 theme following IBCS principles:

- Removes unnecessary chartjunk
- Uses appropriate gridlines (horizontal by default)
- Clean typography
- Minimal axis decoration

### Basic Usage

```{r}
#| label: theme-basic
#| fig-width: 7
#| fig-height: 4
p <- ggplot(mpg, aes(x = class, y = hwy)) +
    geom_boxplot(fill = ibcs_colors()$actual) +
    labs(
        title = "Highway Fuel Economy by Vehicle Class",
        subtitle = "Data from EPA fuel economy tests",
        x = "Vehicle Class",
        y = "Highway MPG"
    )

# Without theme
p
```

```{r}
#| label: theme-applied
#| fig-width: 7
#| fig-height: 4
# With IBCS theme
p + theme_ibcs()
```

### Theme Parameters

```{r}
#| eval: false
theme_ibcs(
    base_size = 11,      # Base font size
    base_family = "",    # Base font family
    grid_x = FALSE,      # Show vertical gridlines
    grid_y = TRUE        # Show horizontal gridlines
)
```

### Customizing Gridlines

```{r}
#| label: theme-grid
#| fig-width: 7
#| fig-height: 4
# Both gridlines
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(color = ibcs_colors()$actual, size = 3) +
    labs(title = "Weight vs Fuel Economy", x = "Weight (1000 lbs)", y = "MPG") +
    theme_ibcs(grid_x = TRUE, grid_y = TRUE)
```

```{r}
#| label: theme-no-grid
#| fig-width: 7
#| fig-height: 4
# No gridlines
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(color = ibcs_colors()$actual, size = 3) +
    labs(title = "Weight vs Fuel Economy", x = "Weight (1000 lbs)", y = "MPG") +
    theme_ibcs(grid_x = FALSE, grid_y = FALSE)
```

### Adjusting Base Size

```{r}
#| label: theme-size
#| fig-width: 7
#| fig-height: 4
# Larger text for presentations
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
    geom_col(stat = "summary", fun = "mean", fill = ibcs_colors()$actual) +
    labs(title = "Average MPG by Cylinders", x = "Cylinders", y = "Mean MPG") +
    theme_ibcs(base_size = 14)
```

## Combining Theme and Colors

```{r}
#| label: combined-example
#| fig-width: 8
#| fig-height: 5
cols <- ibcs_colors()

# Sample data
set.seed(42)
monthly_data <- data.frame(
    month = factor(month.abb[1:6], levels = month.abb[1:6]),
    sales = c(100, 110, 95, 120, 115, 130),
    variance = c(5, 10, -5, 15, 10, 20)
)

monthly_data$var_color <- ifelse(
    monthly_data$variance >= 0,
    cols$variance_positive,
    cols$variance_negative
)

ggplot(monthly_data, aes(x = month)) +
    geom_col(aes(y = sales), fill = cols$actual, width = 0.7) +
    geom_point(
        aes(y = sales, color = var_color),
        size = 4, shape = 18
    ) +
    geom_text(
        aes(y = sales + 5, label = sprintf("%+d", variance), color = var_color),
        fontface = "bold", size = 3.5
    ) +
    scale_color_identity() +
    labs(
        title = "Monthly Sales Performance",
        subtitle = "With variance indicators",
        x = NULL,
        y = "Sales (units)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_ibcs()
```

## Best Practices

### Do
- Use gray tones for primary data representation
- Reserve red/green exclusively for variances
- Keep visualizations clean and uncluttered
- Use consistent colors across all charts in a report

### Don't
- Use rainbow color schemes
- Add unnecessary decorative elements
- Use 3D effects or shadows
- Mix different color conventions in the same report
