---
title: "IBCS Variance Charts"
vignette: >
  %\VignetteIndexEntry{IBCS Variance Charts}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
```

## What are Variance Charts?

Variance charts show the difference between actual values and a reference
(such as plan, budget, or previous year). IBCS recommends using variance
visualization to highlight performance against targets.

### Key Principles

- **Color coding**: Green = favorable, Red = unfavorable
- **Show both absolute and relative variances** when possible
- **Integrated display**: Combine actual values with variance indicators

## Basic Variance Chart

The `ibcs_variance_chart()` function creates IBCS-compliant variance
visualizations:

```{r}
#| label: basic-variance
#| fig-width: 8
#| fig-height: 5
sales_data <- data.frame(
    product = c("Widget A", "Widget B", "Widget C", "Widget D", "Widget E"),
    actual = c(120, 85, 105, 95, 130),
    plan = c(100, 90, 100, 100, 120)
)

ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan
)
```

## Function Parameters

```{r}
#| eval: false
ibcs_variance_chart(
    data,                      # Data frame
    x,                         # Category column
    actual,                    # Actual values column
    reference,                 # Reference values column (plan, PY, etc.)
    show_variance_bars = TRUE, # Show reference as outlined bar
    show_variance_pct = TRUE,  # Show percentage labels
    horizontal = FALSE,        # Horizontal orientation
    sort_by = "none"           # Sort: "none", "actual", or "variance"
)
```

## Horizontal Orientation

For many categories, horizontal bars work better:

```{r}
#| label: horizontal
#| fig-width: 8
#| fig-height: 6
regional_data <- data.frame(
    region = c("North America", "Europe", "Asia Pacific",
               "Latin America", "Middle East", "Africa"),
    actual = c(250, 180, 220, 95, 60, 45),
    plan = c(230, 200, 200, 100, 70, 50)
)

ibcs_variance_chart(
    regional_data,
    x = region,
    actual = actual,
    reference = plan,
    horizontal = TRUE
)
```

## Sorting Options

### Sort by Actual Value

```{r}
#| label: sort-actual
#| fig-width: 8
#| fig-height: 5
ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan,
    sort_by = "actual",
    horizontal = TRUE
) + labs(title = "Products Sorted by Actual Sales")
```

### Sort by Variance

```{r}
#| label: sort-variance
#| fig-width: 8
#| fig-height: 5
ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan,
    sort_by = "variance",
    horizontal = TRUE
) + labs(title = "Products Sorted by Variance (worst to best)")
```

## Customizing the Display

### Without Variance Bars

Show only actual values with variance indicators:

```{r}
#| label: no-bars
#| fig-width: 8
#| fig-height: 4
ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan,
    show_variance_bars = FALSE
) + labs(title = "Simplified View: Actual with Variance Indicators")
```

### Without Percentage Labels

```{r}
#| label: no-pct
#| fig-width: 8
#| fig-height: 4
ibcs_variance_chart(
    sales_data,
    x = product,
    actual = actual,
    reference = plan,
    show_variance_pct = FALSE
) + labs(title = "View Without Percentage Labels")
```

## Adding Titles and Annotations

Since `ibcs_variance_chart()` returns a ggplot object, you can add
layers and customize further:

```{r}
#| label: customized
#| fig-width: 9
#| fig-height: 5
ibcs_variance_chart(
    regional_data,
    x = region,
    actual = actual,
    reference = plan,
    horizontal = TRUE,
    sort_by = "variance"
) +
    labs(
        title = "Regional Sales Performance vs Plan",
        subtitle = "FY 2024 Q3 | Values in $M",
        caption = "Source: Internal Sales Database"
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50")
```

## Understanding the Visualization

The variance chart shows:

1. **Solid bars**: Actual values
2. **Outlined bars**: Reference (plan) values
3. **Diamond markers**: Color-coded variance direction
4. **Percentage labels**: Relative variance to reference

```{r}
#| label: annotated
#| fig-width: 9
#| fig-height: 4
#| echo: false
# Create an annotated example
cols <- ibcs_colors()

example_data <- data.frame(
    category = c("Example"),
    actual = 120,
    reference = 100
)

ggplot(example_data, aes(x = category)) +
    # Reference bar (outlined)
    geom_col(aes(y = reference), fill = NA, color = cols$actual,
             linewidth = 0.8, width = 0.6) +
    # Actual bar (solid)
    geom_col(aes(y = actual), fill = cols$actual, width = 0.4) +
    # Variance indicator
    geom_point(aes(y = actual), color = cols$variance_positive,
               shape = 18, size = 5) +
    # Percentage label
    geom_text(aes(y = actual, label = "+20.0%"),
              color = cols$variance_positive, hjust = -0.3, size = 4) +
    # Annotations
    annotate("segment", x = 0.5, xend = 0.7, y = 100, yend = 100,
             arrow = arrow(length = unit(0.2, "cm"))) +
    annotate("text", x = 0.35, y = 100, label = "Plan\n(outlined)",
             hjust = 1, size = 3) +
    annotate("segment", x = 1.5, xend = 1.3, y = 110, yend = 110,
             arrow = arrow(length = unit(0.2, "cm"))) +
    annotate("text", x = 1.65, y = 110, label = "Actual\n(solid)",
             hjust = 0, size = 3) +
    annotate("segment", x = 1.5, xend = 1.15, y = 120, yend = 120,
             arrow = arrow(length = unit(0.2, "cm"))) +
    annotate("text", x = 1.65, y = 120, label = "Variance\nindicator",
             hjust = 0, size = 3) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Anatomy of a Variance Chart", x = NULL, y = "Value") +
    theme_ibcs()
```

## Building Custom Variance Charts

For more control, you can build variance charts manually using the
scales and theme:

```{r}
#| label: custom-build
#| fig-width: 9
#| fig-height: 5
cols <- ibcs_colors()

# Prepare data
custom_data <- sales_data
custom_data$variance <- custom_data$actual - custom_data$plan
custom_data$variance_pct <- (custom_data$variance / custom_data$plan) * 100
custom_data$var_color <- ifelse(custom_data$variance >= 0,
                                 cols$variance_positive,
                                 cols$variance_negative)

# Build chart
ggplot(custom_data, aes(x = reorder(product, variance))) +
    # Variance bars
    geom_col(aes(y = variance, fill = var_color), width = 0.6) +
    scale_fill_identity() +
    # Zero line
    geom_hline(yintercept = 0, color = cols$actual, linewidth = 0.5) +
    # Value labels
    geom_text(
        aes(y = variance, label = sprintf("%+.0f", variance),
            vjust = ifelse(variance >= 0, -0.5, 1.5)),
        color = custom_data$var_color,
        fontface = "bold"
    ) +
    coord_flip() +
    labs(
        title = "Absolute Variance to Plan",
        subtitle = "Sorted by variance",
        x = NULL, y = "Variance (units)"
    ) +
    theme_ibcs()
```

## Best Practices

1. **Use consistent reference points** - Compare to the same reference
   (plan, PY, etc.) within a report section

2. **Sort meaningfully** - Sort by variance to quickly identify
   best/worst performers

3. **Show context** - Include both absolute values and variances
   when space permits

4. **Color consistently** - Always use green for favorable, red for
   unfavorable (consider color-blind alternatives if needed)

5. **Label clearly** - Ensure the reference (plan, budget, PY) is
   clearly stated in the title or subtitle
