---
title: "IBCS Waterfall Charts"
vignette: >
  %\VignetteIndexEntry{IBCS Waterfall Charts}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
---

```{r}
#| label: setup
#| include: false
library(icbs)
library(ggplot2)
```

## What are Waterfall Charts?

Waterfall charts show how an initial value is affected by a series of
intermediate positive or negative values to arrive at a final value.
They're commonly used for:

- **Income statements**: Revenue → Costs → Profit
- **Variance analysis**: Budget → Changes → Actual
- **Bridge charts**: Year 1 → Drivers → Year 2

## Basic Waterfall Chart

```{r}
#| label: basic-waterfall
#| fig-width: 8
#| fig-height: 5
income_statement <- data.frame(
    item = c("Revenue", "COGS", "Gross Profit", "SG&A", "R&D", "Operating Income"),
    value = c(1000, -400, NA, -250, -100, NA),
    type = c("start", "negative", "subtotal", "negative", "negative", "end")
)

ibcs_waterfall(income_statement, x = item, y = value, type = type)
```

## Function Parameters

```{r}
#| eval: false
ibcs_waterfall(
    data,                      # Data frame
    x,                         # Category column (in order)
    y,                         # Value column
    type = NULL,               # Type column: start/end/positive/negative/subtotal
    fill_positive = NULL,      # Color for positive changes
    fill_negative = NULL,      # Color for negative changes
    fill_total = NULL,         # Color for totals
    show_connectors = TRUE,    # Show connecting lines
    show_values = TRUE         # Show value labels
)
```

## Type Column Values

The `type` column controls how each bar is rendered:

| Type | Behavior |
|------|----------|
| `start` | Begins from zero, sets initial running total |
| `positive` | Adds to running total, floats from previous level |
| `negative` | Subtracts from running total, floats from previous level |
| `subtotal` | Shows current running total, resets baseline to zero |
| `end` | Shows final running total from zero |

## Automatic Type Inference

If no `type` column is provided, the function infers types:
- First row = start
- Last row = end
- Other rows = positive/negative based on sign

```{r}
#| label: auto-infer
#| fig-width: 8
#| fig-height: 4
simple_data <- data.frame(
    step = c("Start", "Add Sales", "Subtract Costs", "Add Other", "End"),
    value = c(100, 50, -30, 10, 130)
)

ibcs_waterfall(simple_data, x = step, y = value) +
    labs(title = "Waterfall with Automatic Type Inference")
```

## Income Statement Example

```{r}
#| label: income-statement
#| fig-width: 10
#| fig-height: 5
pnl <- data.frame(
    line_item = c(
        "Net Revenue",
        "Cost of Sales",
        "Gross Profit",
        "R&D Expense",
        "Sales & Marketing",
        "G&A Expense",
        "Operating Income",
        "Interest Expense",
        "Tax Expense",
        "Net Income"
    ),
    value = c(500, -200, NA, -50, -80, -40, NA, -10, -30, NA),
    type = c(
        "start",
        "negative",
        "subtotal",
        "negative",
        "negative",
        "negative",
        "subtotal",
        "negative",
        "negative",
        "end"
    )
)

ibcs_waterfall(pnl, x = line_item, y = value, type = type) +
    labs(
        title = "Income Statement Waterfall",
        subtitle = "FY 2024 | Values in $M"
    )
```

## Variance Bridge Example

Show what drove the change from one period to another:

```{r}
#| label: variance-bridge
#| fig-width: 9
#| fig-height: 5
bridge <- data.frame(
    driver = c("2023 Revenue", "Volume", "Price", "Mix", "FX Impact", "2024 Revenue"),
    value = c(800, 50, 30, -20, -10, NA),
    type = c("start", "positive", "positive", "negative", "negative", "end")
)

ibcs_waterfall(bridge, x = driver, y = value, type = type) +
    labs(
        title = "Revenue Bridge: 2023 to 2024",
        subtitle = "Key drivers of revenue change"
    )
```

## Customizing Appearance

### Custom Colors

By default, IBCS waterfall charts use gray tones (not red/green) for the
bars themselves. You can customize if needed:

```{r}
#| label: custom-colors
#| fig-width: 8
#| fig-height: 5
cols <- ibcs_colors()

ibcs_waterfall(
    bridge,
    x = driver,
    y = value,
    type = type,
    fill_positive = cols$variance_positive,
    fill_negative = cols$variance_negative,
    fill_total = cols$actual
) +
    labs(title = "Revenue Bridge with Variance Colors")
```

### Without Connectors

```{r}
#| label: no-connectors
#| fig-width: 8
#| fig-height: 4
ibcs_waterfall(
    simple_data,
    x = step,
    y = value,
    show_connectors = FALSE
) +
    labs(title = "Waterfall Without Connector Lines")
```

### Without Value Labels

```{r}
#| label: no-labels
#| fig-width: 8
#| fig-height: 4
ibcs_waterfall(
    simple_data,
    x = step,
    y = value,
    show_values = FALSE
) +
    labs(title = "Waterfall Without Value Labels")
```

## Budget Variance Analysis

```{r}
#| label: budget-variance
#| fig-width: 10
#| fig-height: 5
budget_analysis <- data.frame(
    category = c(
        "Budget",
        "Sales Overperformance",
        "Cost Savings",
        "Unexpected Expenses",
        "Currency Loss",
        "Actual"
    ),
    value = c(1000, 80, 45, -60, -15, NA),
    type = c("start", "positive", "positive", "negative", "negative", "end")
)

ibcs_waterfall(budget_analysis, x = category, y = value, type = type) +
    labs(
        title = "Budget to Actual Variance Analysis",
        subtitle = "Q4 2024 | Values in $K"
    )
```

## Multiple Subtotals

Use subtotals to show intermediate sums:

```{r}
#| label: subtotals
#| fig-width: 11
#| fig-height: 5
detailed_pnl <- data.frame(
    item = c(
        "Product Revenue", "Service Revenue", "Total Revenue",
        "Direct Costs", "Indirect Costs", "Gross Profit",
        "Operating Expenses", "Net Profit"
    ),
    value = c(600, 200, NA, -300, -100, NA, -200, NA),
    type = c(
        "start", "positive", "subtotal",
        "negative", "negative", "subtotal",
        "negative", "end"
    )
)

ibcs_waterfall(detailed_pnl, x = item, y = value, type = type) +
    labs(title = "P&L with Multiple Subtotals")
```

## Combining with Other ggplot2 Elements

Since `ibcs_waterfall()` returns a ggplot object, you can add layers:

```{r}
#| label: combined
#| fig-width: 9
#| fig-height: 5
ibcs_waterfall(bridge, x = driver, y = value, type = type) +
    # Add a reference line
    geom_hline(yintercept = 800, linetype = "dashed",
               color = ibcs_colors()$previous) +
    annotate("text", x = 6, y = 810, label = "2023 Level",
             hjust = 1, size = 3, color = ibcs_colors()$previous) +
    labs(
        title = "Revenue Bridge with Reference Line",
        subtitle = "Comparison to prior year baseline"
    )
```

## Best Practices

1. **Order matters**: Arrange items in a logical flow (e.g., revenue
   before costs, larger items before smaller)

2. **Use subtotals sparingly**: Too many subtotals can clutter the chart

3. **Label clearly**: Ensure each bar's contribution is understandable

4. **Keep it simple**: Waterfall charts work best with 5-10 items;
   aggregate if you have more

5. **Color judiciously**: IBCS recommends gray tones for waterfall bars,
   reserving red/green for explicit variance displays

6. **Show the math**: Values should clearly show the calculation path
   from start to end
